Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-03 ms.tex
adding pdf

\author{R. Angus\altaffilmark{\columbia}
% \altaffilmark{\simons}
        J. R. Davenport\altaffilmark{\ww}
% \altaffilmark{\nsf},
        D. Busazi\altaffilmark{\florida},
        D. Foreman-Mackey\altaffilmark{\cca},
        A. Mann\altaffilmark{\columbia}
% \altaffilmark{\hubble},
        S. Oh\altaffilmark{\princeton},
        D. Kipping\altaffilmark{\columbia}
}

@@

Galactic archaeology and exoplanet populations are two rapidly accelerating
fields of interest within astronomy.
Although seemingly unconnected, these fields are linked by a mutual
requirement for precise stellar parameters.
To galactic archaeologists, ages and elemental abundances are the most
important parameters.
Indeed, most galactic archaeology surveys target exactly these properties.
For exoplaneteers, masses and radii have historically been the most important
stellar parameters for understanding planetary systems.
With a growing number of planet hosts with precise masses and radii, attention
is turning toward other parameters, such as ages, to understand the history
and evolution of these systems.
Age is therefore a fundamental stellar parameter of great interest to at least
two large communities of astronomers.
Unfortunately however, it is a difficult attribute to measure for main sequence
F, G, K, and M stars in the field, in part because low-mass dwarfs do not move
far on the Color-Magnitude diagram (CMD) during their hydrogen burning
lifetimes.
Further, competing stellar evolution models predict different ages for the
same star.
Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.

Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
While rotation periods have previously been measured from starspot-induced
flux modulations for hundreds of stars from the ground, space-based
photometric surveys have opened the door to homogeneous ensemble measures of
stellar rotation, and therefore age.
Using precise light curves available from the TESS mission, we hope to
infer ages for nearly 100,000 main sequence field stars, including a few
hundred planet hosts.
The ages of the planet hosts will allow us to characterize trends in planet
properties over time.

Additionally, as rotation both influences and is influenced {\it by} stellar
magnetic fields, it is inextricably related to stellar activity.
With so many TESS targets being M dwarfs which tend to be particularly active,
understanding the magnetic behavior of these stars has never been more
important.

To enable studies of stellar ages and magnetic activity from rotation periods
with TESS, we propose to:
1. Measure accurate rotation periods for every available TESS FFI and
two-minute cadence target.
2. Determine the origin of the mysterious rotation period bimodality
discovered with Kepler by tracing the rotation period distribution over the
whole sky, and out to further distances utilizing public Gaia data.
3. Measure the star formation history across the sky using a new Bayesian
age-dating system.
4. Further characterize the intrinsic uncertainty in gyrochronology ages using
comoving pairs identified in the first Gaia data release.

@@

One of the most remarkable results from the Kepler rotation period catalog was
the discovery of a bimodal period distribution among field stars by
\citet{mcquillan2013}, which is shown in Figure \ref{fig:davenport}.
While a separate sequence of rapid rotation periods had been known in young
stellar clusters due to lower-mass stars settling on to the angular momentum
main sequence slower \citep[\eg][]{barnes2007}, this new bimodality was
detected from slower rotating M dwarfs (T$_\mathrm{eff} <$ 4000).
The bimodality separated M dwarfs into two populations, those with rotation
periods longer than $\sim$ 20 days, and those with periods between $\sim$ 1
and 20 days.
Follow-up analysis of the Kepler data by \citet{mcquillan2014} found the
period bimodality extended to include K dwarfs.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars, as shown in Figure
\ref{fig:davenport} \citep{davenport2017}.
Two formation scenarios have been proposed to explain the observed period
bimodality.
The first scenario, initially proposed by \citet{mcquillan2013}, is the
rotation period distribution reflects the local star formation history, and
thus the bimodality represents a drop in the star formation rate around 600
Myr ago.
This model is supported by both the extension of the bimodality to earlier
spectral types by \citet{davenport2017}, and also the tentative detection that
the two rotation period populations have distinct proper motion distributions.
However, such a variation in the star formation rate on short timescales has
some tension with independent observational efforts to determine the local
star formation history.
While Color–Magnitude diagram inversions from Hipparcos have suggested a
similarly short timescale variation in star formation of $\sim$ 0.5 Gyr
\citep{hernandez2000}, other studies find slower variations over several
Gyr \citep[\eg][]{cignoni2006}.
Using white dwarf cooling models to infer the local formation history
(“cosmochronology”) also supports higher star formation several Gyr ago, but
can rarely achieve age resolution better than $\sim$ 1 Gyr due to small sample
sizes \citep{tremblay2014}.
The spatial extent of such coherent and localized variations in star formation
history is unknown.
The second scenario to explain this feature is that the period bimodality
occurs due to a previously unknown variation in the spin-down evolution for
low-mass stars.
In this model the star formation history would be continuous over the past
$\sim$ 1 Gyr, and around 600 Myr stars would move quickly through the observed
period minima due to this unknown phase transition or feedback mechanism.
While this model is not predicted by angular momentum loss prescriptions,
rapid transitions in rotation period are observed for stars in young clusters.
Stars move quickly from the rapidly rotating “convective” sequence (periods of
$\sim<1$ day) to the “interface” sequence (periods of several days) during the
first few hundred Myr, with lower mass stars taking longer to make this
transition as they settle onto the main sequence \citep{barnes2003}.
Secondly, a gap in chromospheric activity levels for solar-type field stars
has also been observed \citep{vaughan1980}.
While this magnetic activity indicator smoothly varies with stellar ages over
long timescales, the gap indicates a rapid transition phase from “active” to
“inactive” is present within the first Gyr \citep{pace2009}.
Thirdly, the angular momentum loss underpinning gyrochronology seems to slow
for stars older than the Sun, indicating a potential change in the magnetic
dynamo for slower rotators \citep{angus2015, van-saders2016}.

The TESS FFI and CTL targets will provide the ideal dataset to test these two
formation scenarios.
If the bimodality is due to an age distribution we would expect only to see
the feature locally; it could disappear at greater distances or along
different lines of sight where small scale variations in the star formation
history are less apparent.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, we would expect to find no little to no variation in
this feature as seen in Figure \ref{fig:davenport} with galactic position.

To map the rotation period distribution as a function of galactocentric
position we will match the TESS FFI and CTL targets to the upcoming data
release from the Gaia mission \citep{perryman2001}.
This will also allow us to remove contaminating subgiants and binary stars
from the sample, leaving only single main sequence stars.
The period distribution of G dwarfs is critical for this analysis since they
are observable at greater distances, allowing us to sample different star
formation histories.
With the April 2018 data release from Gaia (DR2) estimate that we will be able
to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$ 3 kpc.
Unlike \Kepler, since \TESS\ is targeting the whole sky, it will be possible
to investigate the rotation period distribution as a function of
galactocentric position.
The size scale over which the Milky Way’s star formation history varies is
unknown however significant variations in star formation histories are seen in
Andromeda between 100 pc volumes, as well as large galaxy-wide trends
\citep[\eg][]{lewis2015}.

@@

Gyrochronology is by far the most widely applicable, relatively precise
age-dating method available.
Although the phenomenon of magnetic braking has been closely studied since its
discovery in the mid/late 20th century, the manifestations of magnetic fields
and their effects on rotation and activity are complex; the physics
underpinning rotational evolution are far from well-understood.
Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
rapidly than the \citet{skumanich1972} law predicts they should.
\citet{van-saders2016} find that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Once a star reaches a critical Rossby number (the ratio of the rotation period
to the convective overturn time), magnetic braking efficiency effectively
drops to zero.
This means a 5 Gyr star of Solar mass will have the same rotation period as a
6 Gyr and rotation period cannot be used as an age proxy.
The transition occurs at around Solar age for Solar mass stars but later for
less massive stars and earlier for more massive stars.
Although this effect reduces the usefulness of gyrochronology for stars more
massive than the Sun, in practise, for Solar-mass stars, down to M dwarfs with
masses greater than 0.35$M_\odot$, gyrochronology can still be considered an
effective dating method.

{\bf We will use the rotation periods of TESS field dwarfs to calibrate the
gyrochronology relations using galactic kinematics}.
Many stars in the TESS CTL will also have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate both
galactocentric positions and action angles of the stars, both of which are age
indicators.
Vertical action (angular momentum through the plane of the Milky Way)
increases over time due to orbital heating and has been demonstrated to
increase with age \citep[\eg][]{west2011}.
Galactocentric position can also determine a star's age, since different
stellar populations, thin or thick disk, halo and bulge have different average
ages.
PI Angus is currently using the \kepler\ rotation periods of stars in the
first \Gaia\ data release to calibrate the relations between rotation period
and vertical action dispersion, two tracers of age.
Since this relation is likely vary with both galactic longitude and latitude,
the rotation periods produced from \TESS\ light curves will enable a
comprehensive, all-sky calibration of these relations.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$.

>>>>

\author{R. Angus\altaffilmark{\columbia}
        J. R. Davenport\altaffilmark{\ww}
        D. Foreman-Mackey\altaffilmark{\cca},
        S. Oh\altaffilmark{\princeton},
        D. Busazi\altaffilmark{\florida},
        A. Mann\altaffilmark{\columbia}
        D. Kipping\altaffilmark{\columbia}
}

@@

Age is a fundamental stellar parameter of great interest to at galactic
archaeologists and exoplaneteers alike but unfortunately it is a difficult
attribute to measure for main sequence F, G, K, and M stars in the field.
Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.

Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
While rotation periods have previously been measured from starspot-induced
flux modulations for hundreds of stars from the ground, space-based
photometric surveys have opened the door to homogeneous ensemble measures of
stellar rotation, and therefore age.
Using precise light curves available from the TESS mission, we expect to
measure the rotation periods of around ... stars.
Many of these rotation periods will be convertible into ages as they will be
intermediate age FGK and early M dwarfs.
Of particular interest will be the expected ~300 gyrochronal ages of exoplanet
host stars.
The gyrochronal ages of these planet hosts, combined with their isochronal and
kinematic ages will allow us to characterize trends in planet properties over
time and across the galaxy.

Additionally, as rotation both influences and is influenced {\it by} stellar
magnetic fields, it is inextricably related to stellar activity.
With so many \TESS\ targets being M dwarfs which tend to be particularly
active, understanding the magnetic behavior of these stars has never been more
important.

We propose to:
1. Measure accurate rotation periods for every available TESS FFI and
two-minute cadence target.
2. Determine the origin of the mysterious rotation period bimodality
discovered with Kepler by tracing the rotation period distribution over the
whole sky, and out to further distances utilizing public Gaia data.
3. Measure the star formation history across the sky using a new Bayesian
age-dating system.
4. Further characterize the intrinsic uncertainty in gyrochronology ages using
comoving pairs identified in the first Gaia data release.
5. Infer the distribution of exoplanets as a function of time and galactic
position.

@@

One of the most remarkable results from the Kepler rotation period catalog was
the discovery of a bimodal period distribution among field stars by
\citet{mcquillan2013}, which is shown in Figure \ref{fig:davenport}.
The bimodality separats M dwarfs into two populations, those with rotation
periods longer than $\sim$ 20 days, and those with periods between $\sim$ 1
and 20 days.
Follow-up analysis of the Kepler data by \citet{mcquillan2014} found the
period bimodality extended to include K dwarfs.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars, as shown in Figure
\ref{fig:davenport} \citep{davenport2017}.
Two scenarios have been proposed to explain the observed period bimodality.
In the first scenario, initially proposed by \citet{mcquillan2013}, the
rotation period distribution reflects the local star formation history, and
thus the bimodality represents a drop in the star formation rate around 600
Myr ago.
This model is supported by both the extension of the bimodality to earlier
spectral types by \citet{davenport2017}, and also the tentative evidence for
two distinct proper motion distributions.
However, such a variation in the star formation rate on short timescales is in
tension with independent observational efforts to determine the local star
formation history.
While color–magnitude diagram inversions from Hipparcos have suggested a
similarly short timescale variation in star formation of $\sim$ 0.5 Gyr
\citep{hernandez2000}, other studies find slower variations over several
Gyr \citep[\eg][]{cignoni2006}.
Using white dwarf cooling models to infer the local formation history
(“cosmochronology”) also supports higher star formation several Gyr ago, but
can rarely achieve age resolution better than $\sim$ 1 Gyr due to small sample
sizes \citep{tremblay2014}.
The spatial extent of such coherent and localized variations in star formation
history is unknown.
The second scenario explaining this feature is a previously unknown variation
in the spin-down evolution for low-mass stars.
In this model, the star formation history would be continuous over the past
$\sim$ 1 Gyr, and around 600 Myr stars would move quickly through the observed
period minima due to some phase transition or feedback mechanism.
While this model is not predicted by angular momentum loss prescriptions,
rapid transitions in rotation period are observed for stars in young clusters.
Stars move quickly from the rapidly rotating “convective” sequence (periods of
$\sim<1$ day) to the “interface” sequence (periods of several days) during the
first few hundred Myr, with lower mass stars taking longer to make this
transition as they settle onto the main sequence \citep{barnes2003}.
Secondly, a gap in chromospheric activity levels for solar-type field stars
has also been observed \citep{vaughan1980}.
While this magnetic activity indicator smoothly varies with stellar ages over
long timescales, the gap indicates a rapid transition phase from “active” to
“inactive” is present within the first Gyr \citep{pace2009}.
Thirdly, the angular momentum loss underpinning gyrochronology seems to slow
for stars older than the Sun, indicating a potential change in the magnetic
dynamo for slower rotators \citep{angus2015, van-saders2016}.

The TESS FFI and CTL targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a age discontinuous age
distribution, the feature should be local; it could
disappear at greater distances or along different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.

To map the rotation period distribution as a function of galactocentric
position we will match the TESS FFI and CTL targets to the upcoming data
release from the Gaia mission \citep{perryman2001}.
This will also allow us to remove contaminating subgiants and binary stars
from the sample, leaving only single main sequence stars.
With the April 2018 data release from Gaia (DR2), we estimate that we will be
able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$ 3
kpc.

@@

Gyrochronology is by far the most widely applicable, relatively precise
age-dating method available.
Although the phenomenon of magnetic braking has been closely studied since its
discovery in the mid/late 20th century, the manifestations of magnetic fields
and their effects on rotation and activity are complex; the physics
underpinning rotational evolution are far from well-understood.
Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
rapidly than the \citet{skumanich1972} law predicts they should.
\citet{van-saders2016} find that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Once a star reaches a critical Rossby number (the ratio of the rotation period
to the convective overturn time), magnetic braking efficiency effectively
drops to zero.
This means a 5 Gyr star of Solar mass will have the same rotation period as a
6 Gyr and rotation period cannot be used as an age proxy.
The transition occurs at around Solar age for Solar mass stars but later for
less massive stars and earlier for more massive stars.
Further calibration is needed however; a lack of rotation periods and
reliable ages for old and low-mass stars leaves a question mark hanging over
the rotational behaviour of stars older than the Sun.
\TESS\ will provide rotation periods for old stars that fall in the continuous
viewing zones.

{\bf We will use the rotation periods of TESS field dwarfs to test the
gyrochronology relations using galactic kinematics}.
Many stars in the TESS CTL will also have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate both
galactocentric positions and action angles of the stars, both of which are age
indicators.
Vertical action (angular momentum through the plane of the Milky Way)
increases over time due to orbital heating and has been demonstrated to
increase with age \citep[\eg][]{west2011}.
Galactocentric position can also determine a star's age, since different
stellar populations, thin or thick disk, halo and bulge have different average
ages.
PI Angus is currently using the \kepler\ rotation periods of stars in the
first \Gaia\ data release to calibrate the relations between rotation period
and vertical action dispersion, two tracers of age.
Since this relation is likely vary with both galactic longitude and latitude,
the rotation periods produced from \TESS\ light curves will enable a
comprehensive, all-sky calibration of these relations.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$.

{\bf We will use the rotation periods of comoving stars identified in the
first \Gaia\ data release to quantify the accuracy and precision of
gyrochronology}.

Although this effect reduces the applicability  of gyrochronology for stars
more massive than the Sun, in practise, for Solar masses and below, down to
0.35$M_\odot$, gyrochronology can still be considered an effective dating
method.
